### 2018-09-11 doc/methods.tex
expanding results section to try and explain variance sources
brms implementation included for funsies

This logistic regression model, minus the final selection of priors, is expressed as 
\begin{equation}
  \begin{aligned}
    t_{i} &\sim \text{Bernoulli}(\Theta) \\
    \Theta_{i} &= \text{logit}^{-1} (X_{i} B_{w[i], p[i]} + a_{d[i], p[i]}) \\
    B_{w, p} &\sim MVN(\alpha_{p}, \Sigma_{B}) \\
    \alpha_{p} &\sim MVN(\mu, \Sigma_{\alpha}) \\
    a_{w, p} &\sim MVN(\delta_{p}, \Sigma_{a}) \\
    \delta_{p} &\sim MVN(0, \Sigma_{\delta}) \\
  \end{aligned}
  \label{eq:core}
\end{equation}
with \(i\) indexing the observation and bracket subscripts referencing the class of the \(i\)th observation where \(w[i]\) is the time of the \(i\)-th observation, \(p[i]\) is the phylum of the \(i\)-th observation, and \(d[i]\) is the age of the \(i\)-th observation.

@@

In total, these prior choices can be expressed as
\begin{equation}
  \begin{aligned}
    \mu_{d} &\sim 
      \begin{cases}
        N(0, 10) & \text{if } d = 1 \\
        N(-1, 3) & \text{if } d = 2 \\
        N(0, 3) & \text{if } d > 2 \\
      \end{cases}
    \delta &\sim N(0, 1) \\
    \Sigma_{B} &= diag(\tau_{B}) \Omega_{B} diag(\tau_{B}) \\
    \Sigma_{\alpha} &= diag(\tau_{\alpha}) \Omega_{\alpha} diag(\tau_{\alpha}) \\
    \Sigma_{a} &= diag(\tau_{a}) \Omega_{a} diag(\tau_{a}) \\
    \Sigma_{\delta} &= diag(\tau_{\delta}) \Omega_{\delta} diag(\tau_{\delta}) \\
    \tau_{B} &\sim C^{+}(5) \\
    \tau_{\alpha} &\sim C^{+}(5) \\
    \tau_{a} &\sim C^{+}(5) \\
    \tau_{\delta} &\sim C^{+}(5) \\
    \Omega_{B} &\sim LKJ(1) \\
    \Omega_{\alpha} &\sim LKJ(1) \\
    \Omega_{a} &\sim LKJ(1) \\
    \Omega_{\delta} &\sim LKJ(1). \\
  \end{aligned}
  \label{eq:priors}
\end{equation}

@@

We implemented our model using the \begin{texttt}rstanarm\end{texttt} package for the R programming language CITATION. This package provides an interface to the Stan probabilistic programming language for writing hierarchical/mixed-effects models in native R. Posterior estimates were obtained through Hamiltonian Monte Carlo, using 2000 steps divided equally between warm-up and sampling. In order to prevent divergent transitions the adapt delta value was increased to 0.9999; all other HMC/NUTS sampling parameters were kept at the defaults for rstanarm version XX CITATION. An example line of R code to fit this mode given a data.frame of all necessary data called ``data'' could be written as:
\begin{verbatim}
  stan_glmer(formula = event ~ range + range_diff + temp + temp_lag + 
                       (1 + range + range_diff + temp + temp_lag | mybin/phylum) + 
                       (1 | age/phylum), 
             prior = normal(c(-1, 0, 0, 0), rep(3, 4), autoscale = FALSE), 
             prior_intercept = normal(0, 10, autoscale = FALSE), 
             prior_aux = cauchy(0, 5, autoscale = FALSE), 
             data = data, 
             adapt_delta = 0.9999, 
             thin = 4)
\end{verbatim}

Posterior convergence was determined using the following general and HMC specific diagnostic criteria: scale reduction factor (\(\hat{R}\); target \(<1.1\)), effective sample size (eff; target value eff/steps \(<0.0001\)), number of samples that saturated the maximum trajectory length for avoiding infinite loops (treedepth; target value 0), sample divergence, and the energy Bayesian Fraction of Mission Information (E-BFMI; target value \(>0.2\)). For further explanation of these diagnostic criteria, see the Stan Manual CITATION.

@@

>>>>

This logistic regression model, minus the final selection of priors, is expressed as 
\begin{equation}
  \begin{aligned}
    t_{i} &\sim \text{Bernoulli}(\Theta) \\
    \Theta_{i} &= \text{logit}^{-1} (X_{i} B_{w[i], p[i]} + a_{d[i], p[i]}) \\
    B_{w, p} &\sim MVN(\alpha_{p}, \Sigma_{B}) \\
    \alpha_{p} &\sim MVN(\mu, \Sigma_{\alpha}) \\
    a_{w, p} &\sim \mathrm{N}(\delta_{p}, \sigma_{a}) \\
    \delta_{p} &\sim \mathrm{N}(0, \sigma_{\delta}) \\
  \end{aligned}
  \label{eq:core}
\end{equation}
with \(i\) indexing the observation and bracket subscripts referencing the class of the \(i\)th observation where \(w[i]\) is the time of the \(i\)-th observation, \(p[i]\) is the phylum of the \(i\)-th observation, and \(d[i]\) is the age of the \(i\)-th observation.

@@

In total, these prior choices can be expressed as
\begin{equation}
  \begin{aligned}
    \mu_{d} &\sim 
      \begin{cases}
        N(0, 10) & \text{if } d = 1 \\
        N(-1, 3) & \text{if } d = 2 \\
        N(0, 3) & \text{if } d > 2 \\
      \end{cases}
    \delta &\sim N(0, 1) \\
    \Sigma_{B} &= diag(\tau_{B}) \Omega_{B} diag(\tau_{B}) \\
    \Sigma_{\alpha} &= diag(\tau_{\alpha}) \Omega_{\alpha} diag(\tau_{\alpha}) \\
    \tau_{B} &\sim C^{+}(5) \\
    \tau_{\alpha} &\sim C^{+}(5) \\
    \Omega_{B} &\sim LKJ(1) \\
    \Omega_{\alpha} &\sim LKJ(1) \\
  \end{aligned}
  \label{eq:priors}
\end{equation}

@@

We implemented our model using the \begin{texttt}rstanarm\end{texttt} package for the R programming language CITATION. This package provides an interface to the Stan probabilistic programming language for writing hierarchical/mixed-effects models in native R. Posterior estimates were obtained through Hamiltonian Monte Carlo, using 2000 steps divided equally between warm-up and sampling. In order to prevent divergent transitions the adapt delta value was increased to 0.9999; all other HMC/NUTS sampling parameters were kept at the defaults for rstanarm version XX CITATION.

An implementation of our full model in \begin{texttt}rstanarm\end{texttt}, given a data.frame of all necessary data in a data.frame called ``data'', is coded as
\begin{verbatim}
  stan_glmer(formula = event ~ range + range_diff + temp + temp_lag + 
                       (1 + range + range_diff + temp + temp_lag | mybin/phylum) + 
                       (1 | age/phylum), 
             data = data, 
             family = 'binomial',
             prior = normal(c(-1, 0, 0, 0), rep(3, 4), autoscale = FALSE), 
             prior_intercept = normal(0, 10, autoscale = FALSE), 
             prior_aux = cauchy(0, 3, autoscale = FALSE), 
             chains = 4,
             thin = 4,
             adapt_delta = 0.9999).
\end{verbatim}

Similarly, our full model can be implemented in \begin{texttt}brms\end{texttt} as
\begin{verbatim}
priors <- c(set_prior('normal(0, 10)', class= 'Intercept'),
            set_prior('normal(0, 1)', class = 'b'),
            set_prior('normal(-1, 1)', class = 'b', coef = 'range'),
            set_prior('cauchy(0, 3)', class = 'sd'),
            set_prior('lkj(1)', class = 'cor'))
brmfit <- brm(formula = bf(event ~ range + range_diff + temp + temp_lag +
                           (1 + range + range_diff + temp + temp_lag | mybin/phylum) +
                           (1 | age/phylum))
              data = data, 
              family = bernoulli(), 
              prior = priors,
              chains = 4, 
              thin = 4,
              control = list(adapt_delta = 0.999999).
\end{verbatim}

Posterior convergence was determined using the general and HMC-specific diagnostic criteria: scale reduction factor (\(\hat{R}\); target \(<1.1\)), effective sample size (eff; target value eff/steps \(<0.0001\)), number of samples that saturated the maximum trajectory length for avoiding infinite loops (treedepth; target value 0), sample divergence, and the energy Bayesian Fraction of Mission Information (E-BFMI; target value \(>0.2\)). For further explanation of these diagnostic criteria, see the Stan Manual CITATION.

@@

%\section{Variance Partioning and Intraclass Correlation Coeficient}
%
%There are multiple levels and sources of variance in this model. Each of the grouping factors account for some amount of variance in the data. Normalizing that variance by the total variance from all sources, we can see the percent contribution to variance. 
%
%``level 1'' is the ``data level.'' in a regular linear regression problem, this would be the residual variance. that's not calculated in logistic regression, which poses a problem. how do we get this value? simulation?
%
%``level 2'' is the ``time level.'' the logistic regression coefficients \textbf{vary} through time, we are actually estimating that variance in \(\Sigma_{B}\) and \(\tau_{B}\). there is also the non-nested age-based varying-intercept. this source of variance is in \(\Sigma_{a}\) and \(\tau_{a}\).
%
%``level 3'' is the ``phyla level.'' the over-time averages of the regression coefficients further vary between phyla. that variance is part of \(\Sigma_{\alpha}\) and \(\tau_{\alpha}\). there is also the average age effect on extinction probability which is part of \(\Sigma_{\delta}\) and \(\tau_{\delta}\).
%
%that's a lot of variance. 
%
%we want to figure out what the total variance is, so we can determine the percent contribution for each level and segment of variance. the level 1 issue and the nonnestedness pose potential challenges.
